\textsf{Are large language models superhuman chemists?}

### 2024-10-29 src/tex/appendix.tex
fix changes from tiny to chembenchmini

@@

\clearpage
\subsection{Summary of trends and recommendations}

\oone consistently leads across both \ChemBench corpus and \chembenchmini.  \ClaudeThreeFiveSonnet, \GPTFourO, \LlamaThreeOneFourZeroFiveBInstruct form the next tier of strong performers. These models show robust performance across all skills and difficulty levels.

Larger models generally perform better (\LlamaThreeOneFourZeroFiveBInstruct > \LlamaThreeOneSeventyBInstruct > \LlamaThreeOneEightBInstruct). However, smaller but well-designed models can compete with larger models. For example, \GemmaTwoNineBIt has an overall accuracy of 0.48 on \ChemBench corpus, which is only 10\% less accurate than the much larger \LlamaThreeOneFourZeroFiveBInstruct model.

One can observe a clear progression of performance across model families (e.g., \ClaudeThreeFiveSonnet >  \ClaudeThree > \ClaudeTwo or \GPTFourO > \GPTFour > \GPTThreeFiveTurboZeroT). Newer versions consistently outperform their predecessors, indicating improvements in scientific understanding of these model families.
Scores on knowledge-intensive tasks are typically lower than those on calculation and reasoning-intensive questions. Performance on reasoning-intensive questions in \chembench tends to be the highest.

Based on these findings, some scope for improving these models could be to focus on enhancing knowledge-based training, possibly through improved pre-training on chemistry-specific texts or with chemistry knowledge-base integration. However, this type of knowledge should be integrated with sound machine-learning approaches. For example, \GalacticaOneTwentyB showed the lowest performance in our study, even though it was specifically trained on scientific text. This suggests that the knowledge embedded in papers and textbooks has not been fully leveraged.

Models show the strongest performance in reasoning tasks on the \chembench corpus, which suggests that current training approaches are good at developing logical/reasoning capabilities for basic problem-solving of university-level exams. However, performance drops significantly for advanced tasks, suggesting that more advanced chemistry problems with complex, multi-step solutions should be included in training or finetuning.

\clearpage
\subsection{Leaderboard}
Our leaderboard is based on the tool chain developed for Matbench.\autocite{Dunn_2020}
Briefly, the \chembench pipeline produces standardized files in \texttt{json} format that contributors can add via pull requests to the \chembench repository.
The Markdown tables and interactive plots are automatically generated and updated on the \chembench website. The leaderboard is available at \url{https://lamalab-org.github.io/chem-bench/leaderboard/}.

>>>>

\begin{table}
    \caption{\textbf{Performance of the models on the \chembench corpus.} The table shows the fraction of questions answered correctly by the models for different skills and difficulty levels.}
    \resizebox{\textwidth}{!}{
    \variable{output/performance_table.tex}
    }
    \label{tab:performance_table}
\end{table}

\begin{table}
    \caption{\textbf{Performance of the models on \chembenchmini.} The table shows the fraction of questions answered correctly by the models for different skills and difficulty levels.}
    \resizebox{\textwidth}{!}{
    \variable{output/performance_table_human_subset.tex}
    }
    \label{tab:performance_table_human_subset}
\end{table}

\begin{figure}[htb]
    \centering
    \includegraphics[width=\textwidth]{figures/all_questions_models_completely_correct_radar_overall.pdf}
    \caption{\textbf{Performance of the models on the different topics of the \chembench corpus.} The radar plot shows the performance of the models on the different topics of the \chembench corpus. The performance is measured as the fraction of questions answered correctly by the models.
    A score of 1 (full coverage until the outer line of this plot) indicates that all questions were answered correctly, while a score of 0 indicates that none were answered correctly.
    }
    \label{fig:all_questions_models_completely_correct_radar_overall}
    \script{analyze_model_reports.py}
\end{figure}

\begin{figure}[htb]
    \centering
    \includegraphics{figures/performance_per_topic.pdf}
    \caption{\textbf{Fraction of correctly answered questions per data source on \chembenchmini.} The heatmap shows, in color, the fraction of questions answered correctly by different systems for some of our data sources. The performance is measured as the fraction of questions answered correctly by the models. A score of one (red) indicates that all questions were answered correctly, while a score of zero (blue) indicates that none of the questions were answered correctly.
        We see that the performance of the models varies significantly between the different data sources. For instance, it is interesting to observe that questions sourced based on textbooks seem easier for our leading models than for humans. However, this performance does not correlate with performance on other sources, e.g., semi-programmatically created tasks such as questions about the number of signals in an \gls{nmr} spectrum.
    }
    \label{fig:performance_per_topic}
    \script{analyze_performance_per_source.py}
\end{figure}

\Cref{fig:performance_corpus_and_tiny} shows the performance of the models on the \chembench corpus and the \chembenchmini subset. The relative performance difference between both is quite similar across most models. 
This makes \chembenchmini subset a reliable subset for human baseline comparison and particularly valuable for rapid prototyping and initial model assessment phases.

An interesting observation is the significant impact of chemical preference tasks on \GPTFour's scores. A detailed breakdown of overall accuracy into scores on different skills and difficulty levels is provided in \Cref{table:performance_table} and \Cref{tab:performance_table_human_subset} for \chembench corpus and the \chembenchmini subset, respectively.

\begin{figure}[htb]
    \centering
    \includegraphics{figures/performance_per_topic_tiny.pdf}
    \caption{\textbf{Fraction of correctly answered questions per data source on \chembenchmini.} The heatmap shows, in color, the fraction of questions answered correctly by different systems for some of our data sources. The performance is measured as the fraction of questions answered correctly by the models. A score of one (red) indicates that all questions were answered correctly, while a score of zero (blue) indicates that none were answered correctly.
        We see that the performance of the models varies significantly between the different data sources. For instance, it is interesting to observe that questions sourced based on textbooks seem easier for the leading models than for humans. However, this performance does not correlate with performance on other sources, e.g., semi-programmatically created tasks such as questions about the number of signals in an \gls{nmr} spectrum.
    }
    \label{fig:performance_per_topic_tiny}
    \script{analyze_performance_per_source.py}
\end{figure}

\begin{figure}[htb]
    \centering
    \includegraphics{figures/corpus_human_comparison.pdf}
    \caption{\textbf{
        Performance of the models on the \chembench corpus and \chembenchmini subset.} The bar plot shows the fraction of questions that were answered completely correctly, highlighting the relative performance of different models on both the \chembench corpus and the \chembenchmini subset.
        We see that the model ranking remains fairly consistent across both sets}  
    \label{fig:performance_corpus_and_tiny}
    \script{corpus_humanset_performance.py}
\end{figure}

@@

\clearpage
\subsection{Leaderboard}
\label{sec:leaderboard}
Our leaderboard is based on the tool chain developed for Matbench.\autocite{Dunn_2020}
Briefly, the \chembench pipeline produces standardized files in \texttt{json} format that contributors can add via pull requests to the \chembench repository.
The Markdown tables and interactive plots are automatically generated and updated on the \chembench website. The leaderboard is available at \url{https://lamalab-org.github.io/chem-bench/leaderboard/}.
